\textbf{Betelwise1: A Hybrid Conventional-CNN Approach for Efficient Planetary Seismic Event Detection and Analysis}

### 2025-05-24 paper.tex
updated

Machine learning (ML), particularly deep learning techniques like Convolutional Neural Networks (CNNs), has emerged as a powerful tool for analyzing complex patterns in geophysical data \cite{b16, b17}. In planetary seismology, CNNs have shown promise for tasks such as detecting moonquakes from Apollo data. For instance, Civilini et al. (2021) demonstrated the use of CNNs trained on spectrograms of Earth-based seismic events, employing transfer learning to successfully identify moonquakes in the Apollo PSE and LSPE datasets \cite{b1}. Their work highlighted the potential of ML to overcome the limitations of local data scarcity by leveraging non-local training sets.

@@

\subsection{Machine Learning in Planetary Seismology}
The application of ML to planetary seismology is a more recent but rapidly growing field, driven by the unique challenges of these missions. The work by Civilini et al. (2021) \cite{b1} is a significant contribution in this area. They developed CNN models trained on spectrograms of seismic events from a single Earth station and successfully applied these, using transfer learning, to detect moonquakes in the Apollo Passive Seismic Experiment (PSE) and Lunar Seismic Profiling Experiment (LSPE) data. Their approach demonstrated the feasibility of using non-local training data and highlighted the potential for CNNs to catalog planetary seismicity even without prior local seismic data. They also introduced an "extra-arrival accuracy metric" to quantify performance on noisy lunar datasets.

@@

Our Asteroseismology framework builds upon these foundations. It acknowledges the strengths of conventional algorithms like STA/LTA for initial, computationally cheap candidate generation, similar to some multi-stage earthquake detection systems on Earth. However, it significantly differs from approaches like Civilini et al. \cite{b1} by using a hybrid pipeline that feeds 1D waveform segments and engineered auxiliary features directly to the CNN, rather than relying solely on spectrograms. This aims to retain fine temporal details and leverage interpretable features, potentially enhancing both efficiency and robustness for onboard processing.

@@

Compared to purely conventional methods, Asteroseismology offers substantially lower false positive rates. When compared to ML approaches like that of Civilini et al. \cite{b1}, which primarily used spectrograms as CNN input, our use of 1D waveforms combined with auxiliary statistical features offers a different pathway for feature extraction. While a direct quantitative comparison is complex without reimplementing their specific model on our exact data splits, our approach aims to preserve fine temporal details in the waveform and provide explicit, interpretable auxiliary features to the CNN, which could be advantageous for certain event types or computational constraints. The processing of data from both lunar (ALSEP) and Martian (InSight) missions in our training and testing pipeline also demonstrates a broader applicability than studies focused solely on one body.

@@

\bibitem{b1} F. Civilini, R.C. Weber, Z. Jiang, D. Phillips, and W. David Pan, ``Detecting moonquakes using convolutional neural networks, a non-local training set, and transfer learning,'' \emph{Geophys. J. Int.}, vol. 225, no. 3, pp. 2120--2134, Mar. 2021.

>>>>

Machine learning (ML), particularly deep learning techniques like Convolutional Neural Networks (CNNs), 
has emerged as a powerful tool for analyzing complex patterns in geophysical data \cite{b16, b17}. In 
planetary seismology, CNNs have shown promise for tasks such as detecting moonquakes from Apollo data. 
For instance, Civilini et al. (2021) demonstrated the use of CNNs trained on spectrograms of Earth-based 
seismic events, employing transfer learning to successfully identify moonquakes in the Apollo PSE and 
LSPE datasets \cite{favPaper}. Their work highlighted the potential of ML to overcome the limitations of local 
data scarcity by leveraging non-local training sets.

@@

\subsection{Machine Learning in Planetary Seismology}
The application of ML to planetary seismology is a more recent but rapidly growing field, driven by the unique challenges of these missions. The work by Civilini et al. (2021) \cite{favPaper} is a significant contribution in this area. They developed CNN models trained on spectrograms of seismic events from a single Earth station and successfully applied these, using transfer learning, to detect moonquakes in the Apollo Passive Seismic Experiment (PSE) and Lunar Seismic Profiling Experiment (LSPE) data. Their approach demonstrated the feasibility of using non-local training data and highlighted the potential for CNNs to catalog planetary seismicity even without prior local seismic data. They also introduced an "extra-arrival accuracy metric" to quantify performance on noisy lunar datasets.

@@

Our Asteroseismology framework builds upon these foundations. It acknowledges the strengths of conventional algorithms like STA/LTA for initial, computationally cheap candidate generation, similar to some multi-stage earthquake detection systems on Earth. However, it significantly differs from approaches like Civilini et al. \cite{favPaper} by using a hybrid pipeline that feeds 1D waveform segments and engineered auxiliary features directly to the CNN, rather than relying solely on spectrograms. This aims to retain fine temporal details and leverage interpretable features, potentially enhancing both efficiency and robustness for onboard processing.

@@

Compared to purely conventional methods, Asteroseismology offers substantially lower false positive rates. When compared to ML approaches like that of Civilini et al. \cite{favPaper}, which primarily used spectrograms as CNN input, our use of 1D waveforms combined with auxiliary statistical features offers a different pathway for feature extraction. While a direct quantitative comparison is complex without reimplementing their specific model on our exact data splits, our approach aims to preserve fine temporal details in the waveform and provide explicit, interpretable auxiliary features to the CNN, which could be advantageous for certain event types or computational constraints. The processing of data from both lunar (ALSEP) and Martian (InSight) missions in our training and testing pipeline also demonstrates a broader applicability than studies focused solely on one body.

@@

\bibitem{favPaper} F. Civilini, R.C. Weber, Z. Jiang, D. Phillips, and W. David Pan, ``Detecting moonquakes using convolutional neural networks, a non-local training set, and transfer learning,'' \emph{Geophys. J. Int.}, vol. 225, no. 3, pp. 2120--2134, Mar. 2021.
